Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-03-04 paper/paper.tex
done section 3 (for now)

\section{Illustrative Example}
\label{sec:suri-metas-coverage}
\label{sec:active-recon}

This section briefly shows how \tname{} synthesizes rules for one
example attack. Active Reconnaissance is typically used as a
preliminary step towards an actual attack. It is a method to collect
information about a computing system to determine potential
vulnerabilities. For example, Ping Scan scans the network for IPs of
accessible hosts. An attacker sends several ICMP Echo requests to a
range of IP addresses in the network and checks which ones respond to
these requests. This method enables an attacker to discover which
machines are active in the network.

@@

Figure~\ref{fig:pingscan-example} shows a Suricata rule that captures
a Ping Scan, as documented in NMAP~\cite{netmap}, a network discovery
and security auditing open-source tool. The essential options to
capture the attack (see Section~\ref{sec:example-suricata-rules}) are
in bold---the option \CodeIn{dsize: 0} indicates that the packet has
no bytes in the payload and the option \CodeIn{itype: 8} indicates
that the packet to be captured are ICMP Echo requests.

In the following, we briefly show how \tname{} synthesizes rules for
this attack. The action is left constant (\CodeIn{alert}) and the
header is inferred from the protocol. For example, in this case,
\tname{} sets the header to \CodeIn{icmp any any -> any any},
indicating that the rule applies to \CodeIn{icmp} messages and applies
to the most general flow description, \ie{}, it instructs the
\nids\ to analyze the traffic flowing from/to any address and port.
The rationale for this decision is that the choice of these parameters
are domain-specific and subject to change by the system administrator.

The main challenge in synthesizing rules is to discover the set of
rule options that capture the malicious traffic, but allow the benign
traffic to pass. \tname{} proceeds as follows to synthesize rule
options. First, it analyzes the malicious traffic to extract options
expressed in that traffic. Recall from
Section~\ref{sec:rules-and-packets} that some options can be inferred
from packet contents. \tname{} finds the following five options at
this stage: {\scriptsize{\texttt{\textbf{dsize}:0; \textbf{itype:}8;
      \textbf{icode:}0; \textbf{icmp\_id:}23570;
      \textbf{icmp\_seq:}3439;}}}.  By construction, a rule that uses
this set of options captures the malicious traffic. Unfortunately, the
rule above is overspecified. Informally, the rule includes ``too
specific'' characteristics of the malicious traffic and could
potentially result in false negatives. In this case, although the
options \CodeIn{dsize:0} and \CodeIn{itype:8} are present in the set
of inferred options (as the golden rule from
Figure~\ref{fig:pingscan-example}), the others are not.

@@

Second, \tname{} uses the benign traffic to mitigate the overfitting
problem. In this step, \tname{} searches for alternative rules that
preserve the invariant to capture the positive malicious traffic but
not capture the negative benign traffic. It uses the rule obtained in
the previous step as seed to bootstrap the search for alternative
plausible rules, \ie{}, rules that satisfy the aforementioned
invariant. The search is designed to generate rules containing subsets
of the options declared in the seed rule. Considering this example,
\tname{} produces a total of \pingscanplausible{}\Fix{check!}
plausible rules at the end of the search.  Note that there are $2^n$
possible plausible rules, where $n$ is the number of options contained
in the seed rule ($n=5$, in this case). \tname{} is able to prune the
search space and discard many implausible rules. It is worth noting
that the number of plausible rules that \tname\ produces at this step
can be very high for other example attacks (see column ``\#~Plausible
Rules'' on Table~\ref{table:results}).

Finally, \tname{} ranks plausible rules according to heuristics
obtained by analyzing a database of existing rules. The general
intuition is that new rules should share characteristics with
pre-existing rules. For example, one heuristic function used by
\tname{} leverages the \emph{size of a rule} for ranking.  The score
of a rule $r$ that this function produces is proportional to the
popularity of the number of options in
$r$. Figure~\ref{fig:distribution-rule-size} shows the distribution of
the number of content options associated with rules from a public data
set~\cite{emerging-threats-open}. Note that rules with 2 to 6 options
are more common. At the end of this step, \tname{} ranks the golden
rule at the top of the ranking, including \pingscanplausible{}
rules. It is worth noting that \tname\ consistently ranks the golden
rule very high in most of the cases even when the technique produces a
large number of plausible rules (see Table~\ref{table:results}).

@@

  \caption{\label{fig:distribution-strings}Options.}

@@

\caption{Distribution of features.}

>>>>

\section{Illustrative Example}
\label{sec:suri-metas-coverage}
\label{sec-:active-recon}

This section briefly shows how \tname{} synthesizes rules for one
example attack. Active Reconnaissance is typically used as a
preliminary step towards an actual attack. It is a method to collect
information about a computing system to determine potential
vulnerabilities. For example, the Ping Scan attack~\cite{ping-scan}
scans the network for IPs of accessible hosts. An attacker sends
several ICMP Echo requests to a range of IP addresses in the network
and checks which ones respond to these requests. This method enables
an attacker to discover which machines are active in the network.

@@

Figure~\ref{fig:pingscan-example} shows a Suricata rule that captures
a Ping Scan, as documented in NMAP~\cite{netmap}, a network discovery
and security auditing open-source tool. The essential options to
capture the attack are in bold---the option \CodeIn{dsize: 0}
indicates that the packet has no bytes in the payload and the option
\CodeIn{itype: 8} indicates that the packet to be captured are ICMP
Echo requests.

In the following, we briefly show how \tname{} synthesizes rules for
this attack. In short, \tname{} takes one or more examples of an
attack---in the form of network messages---and produces rules to
capture that attack. Considering the rule for the Ping Scan attack,
whose example was reproduced with NMAP, \tname{} leaves the action of
the corresponding rule constant (as \CodeIn{alert}) and infers the
header by checking the protocol of the input messages. For example, in
this case, \tname{} sets the header to \CodeIn{icmp any any -> any
  any}, indicating that the rule applies to \CodeIn{icmp} messages and
the most general flow description, \ie{}, it instructs the \nids\ to
analyze the icmp traffic flowing from/to any address and port.  The
rationale for this decision is that the choice of these parameters are
domain-specific and subject to change by the system administrator.

The main challenge in synthesizing rules is to discover the set of
rule options (see Table~\ref{table:rules}) that captures the malicious
traffic, but allows the benign traffic to pass. \tname{} proceeds as
follows to synthesize rule options. First, it analyzes the malicious
traffic to extract options expressed in that traffic. Recall from
Section~\ref{sec:rules-and-packets} that some options can be inferred
from packet contents. \tname{} finds the following five options at
this stage: {\scriptsize{\texttt{\textbf{dsize}:0; \textbf{itype:}8;
      \textbf{icode:}0; \textbf{icmp\_id:}23570;
      \textbf{icmp\_seq:}3439;}}}.  By construction, a rule that uses
this set of options captures the malicious traffic. Unfortunately, the
rule above is overspecified. Informally, the rule includes ``too
specific'' characteristics of the malicious traffic and could
potentially result in false negatives. In this case, although the
options \CodeIn{dsize:0} and \CodeIn{itype:8} are present in the set
of inferred options---as the golden rule from
Figure~\ref{fig:pingscan-example} shows---the others are not.

@@

Second, \tname{} uses the benign traffic to mitigate the overfitting
problem. In this step, \tname{} searches for alternative rules that
preserve the invariant to capture the positive (/malicious) traffic
but not capture the negative benign traffic. It uses the rule obtained
in the previous step as seed to bootstrap the search for alternative
plausible rules, \ie{}, rules that satisfy the aforementioned
invariant. The search is designed to generate rules containing subsets
of the options declared in the seed rule. Considering this example,
\tname{} produces a total of \pingscanplausible{} plausible rules at
the end of the search. This is a smaller number when compared to $2^n$
option combinations, where $n$ is the number of options contained in
the seed rule ($n=5$, in this case). \tname{} is able to prune the
search space and discard many implausible rules. It is worth noting
that the number of plausible rules that \tname\ produces at this step
can be very high for other example attacks (see column ``\#~Plausible
Rules'' on Table~\ref{table:results}).

Finally, \tname{} ranks plausible rules according to heuristics
obtained by analyzing a database of existing rules. The general
intuition is that new rules should share characteristics with existing
human-written rules. For example, one heuristic function used by
\tname{} leverages the popularity of options for ranking.  The score
of a rule $r$, according to that heuristic alone, is proportional to
the popularity of the options declared in
$r$. Figure~\ref{fig:distribution-options} shows the distribution of
options associated with rules from a public
ruleset~\cite{emerging-threats-open} containing
\numrulessuri\ rules. Note that rules with options \CodeIn{content}
and \CodeIn{flow} are the most common. The options \CodeIn{dsize} and
\CodeIn{itype}, the ones associated with the golden rule, appear as
the 6th and 7th most popular, respectively. But, among the options
inferred by \tname{} at the first step, \CodeIn{dsize} and
\CodeIn{itype} are the most popular. At the end of this step, \tname{}
ranks the golden rule second in the list including
\pingscanplausible{} rules. It is worth noting that
\tname\ consistently ranks the golden rule very high in most of the
cases even when the technique produces a large number of plausible
rules (see Table~\ref{table:results}).

@@

  \caption{\label{fig:distribution-options}Options.}

@@

\caption{Distributions.}
